---
title: "Quering SQLite Relational Database"
output: html_notebook
---

## Setup

Creating a connection variable to the SQlite databse to perform all the query operation


```{r}
library(RSQLite)

#provide path to database folder

fpath = 

#provied path to database file

dbfile = "/Users/jayaramanvenkatesan/Desktop/CS5200/SQLite - R - Queries/MediaDB.db"

dbcon <- dbConnect(RSQLite :: SQLite() , dbfile)

#tables <- dbListTables(dbcon)

# Print the table names
#print(tables)


```



## Question 1 :
What are the last name, title, and hire date of all employees?

```{sql connection = dbcon}
Select LastName , Title , HireDate from employees

```


## Question 2 :
What are the titles of each album and the number of distinct genres on each album? 

```{sql connection = dbcon}
SELECT a.Title AS title, COUNT(DISTINCT b.GenreId) AS distinct_genre
  FROM albums AS a
  JOIN tracks AS b ON a.AlbumId = b.AlbumId
  GROUP BY b.AlbumId

```



## Question 3 :
List the number of tracks by genre, ordered from most to least, restricted to those genres for which there are at least three tracks.

```{sql connection = dbcon}
SELECT a.Name AS genre, COUNT(1) AS num_of_tracks
  FROM genres AS a
  JOIN tracks AS b ON a.GenreId = b.GenreId
  GROUP BY b.GenreId
  HAVING COUNT(1) >= 3
  ORDER BY num_of_tracks DESC

```
## Question 4 :
What is the average "age" of the employees in years. 
```{sql connection = dbcon}
SELECT AVG((strftime('%Y', 'now') - strftime('%Y', BirthDate)))  AS avg_age
FROM employees;

```
## Question 5 :
What are the total number of unique purchases for each Brazilian state (e.g., MG, SP, etc.), ordered alphabetically by state?
```{sql connection = dbcon}

select a.State, count(distinct a.LastName) as count 
from customers a 
left join invoices b 
on a.CustomerId = b.CustomerId 
where a.Country = 'Brazil' 
group by a.State 
order by a.State

```


## Question 6 :
How many employees never supported a single customer, i.e., they appear in the employees table but not in the customers table?
```{sql connection = dbcon}
SELECT COUNT(*) AS non_supported_employee
FROM employees a
LEFT JOIN customers b ON a.EmployeeId = b.SupportRepId
WHERE b.SupportRepId IS NULL;
```


## Question 7 :
How many albums contain the substring "Rock" but not "Hard" in the album title?
```{sql connection = dbcon}
SELECT COUNT(*) AS count
FROM albums
WHERE title LIKE '%Rock%' AND title NOT LIKE '%Hard%';

```


## Question 8 :
What is the total "time" of all tracks per media type in hours, i.e., add up the "Milliseconds" column in tracks? Display the media type and the time in hours rounded to one decimals. Only consider media types with at least one track.

```{sql connection = dbcon}
SELECT mt.Name AS MediaType, ROUND(SUM(t.Milliseconds) / (1000.0 * 60 * 60), 1) AS Time_in_hours
FROM tracks AS t
JOIN media_types AS mt ON t.MediaTypeId = mt.MediaTypeId
GROUP BY mt.MediaTypeId

```


## Question 9( WITH RANK) :
Which genre has the most tracks?

```{sql connection = dbcon}
SELECT genre, num_of_tracks
FROM (
  SELECT a.Name AS genre, COUNT(1) AS num_of_tracks, RANK() OVER (ORDER BY COUNT(1) DESC) AS rank
  FROM genres AS a
  JOIN tracks AS b ON a.GenreId = b.GenreId
  GROUP BY b.GenreId
) AS subquery
WHERE rank = 1;

```
## Question 9 ( ALTERNATE WITH LIMIT ) :
Which genre has the most tracks?

```{sql connection = dbcon}
SELECT a.Name AS genre, COUNT(1) AS num_of_tracks
  FROM genres AS a
  JOIN tracks AS b ON a.GenreId = b.GenreId
  GROUP BY b.GenreId
  ORDER BY num_of_tracks DESC limit 1

```



## Question 10

Using a SQL query (from R), retrieve the "GenreId", "Name" (or a genre), and "UnitPrice" (of a track) for all tracks and store the result in a dataframe. Using the data in the dataframe and R code, calculate the mean (µ) and standard deviation (σ) of "UnitPrice" and then calculate the 95% confidence interval for the mean, defined as: µ ± 1.96σ. Display the result in the following format (but not that color and font):

```{r}

query <- "SELECT t.GenreId, g.Name, t.UnitPrice
         FROM tracks AS t FULL JOIN genres g ON g.GenreId = t.GenreId";


df <- dbGetQuery(dbcon, query)

mean_price <- mean(df$UnitPrice)
sd_price <- sd(df$UnitPrice)


lower_bound = mean_price - (1.96 * sd_price)
upper_bound = mean_price + (1.96 * sd_price)
cat("The 95% CI for the mean unit price ranges from", lower_bound , "to", upper_bound,".")
```


