---
title: "Assignment / Explore Query Planning and Indexing"
date : "Summer Full 2023"
output:
  pdf_document: default
  html_notebook: default
---

## Setup

Creating a connection variable to the SQlite databse to perform all the query operation


```{r}
library(RSQLite)

#provide path to database folder

  
#provied path to database file

dbfile = "sakila.db"

dbcon <- dbConnect(RSQLite :: SQLite() , dbfile)

tables <- dbListTables(dbcon)

# Print the table names
print(tables)


```


## Ques 1

Below is the R chunk to perform deletion of all user defined indexes and query 
to find the number of films per language
```{r}

# TODO : Delete user defined indexes

drop_ind <- "DROP INDEX IF EXISTS TitleIndex"

dbExecute(dbcon , drop_ind)

query <- "SELECT l.name AS language_name , COUNT(*) AS Total_films
FROM film as f
JOIN language as l ON f.language_id = l.language_id
GROUP BY l.name"


df <- dbGetQuery(dbcon, query)

print(df)


```


## Ques 2

Below is the R chunk to find the query plan for above query
```{r}


query2  <- " EXPLAIN QUERY PLAN 
SELECT l.name AS language_name , COUNT(*) AS Total_films
FROM film as f
JOIN language as l ON f.language_id = l.language_id
GROUP BY l.name"


result2 <- dbGetQuery(dbcon, query2)

print(result2)

```

## Ques 3

Below is the R chunk to find the title , category name and length of the film,
titled "ZORRO ARK"
```{r}

query3  <- "SELECT f.title , f.length , c.name
  FROM film as f
  JOIN film_category as fc ON f.film_id = fc.film_id
  JOIN category as c ON fc.category_id = c.category_id 
  where f.title = 'ZORRO ARK'
  "

# calculating time taken here for the purpose of ques 8
s_t <- system.time ({
  result3 <- dbGetQuery(dbcon, query3)
})



print(s_t)



print(result3)

```

## Ques 4

Below is the R chunk to explain the query plan of above query
```{r}



query4  <- " EXPLAIN QUERY PLAN
 SELECT f.title , f.length , c.name
  FROM film as f
  JOIN film_category as fc ON f.film_id = fc.film_id
  JOIN category as c ON fc.category_id = c.category_id 
  where f.title = 'ZORRO ARK'
  "


result4 <- dbGetQuery(dbcon, query4)

print(result4)

```

## Ques 5

Below is the R chunk to create index film table
```{r}

dbExecute(dbcon, "CREATE INDEX TitleIndex ON FILM (TITLE)")


```

## Ques 6

Below is the R chunk to create index film table
```{r}

query6_plan  <- " EXPLAIN QUERY PLAN
 SELECT f.title , f.length , c.name
  FROM film as f
  JOIN film_category as fc ON f.film_id = fc.film_id
  JOIN category as c ON fc.category_id = c.category_id 
  where f.title = 'ZORRO ARK'
  "


result6_plan <- dbGetQuery(dbcon, query6_plan)

print(result6_plan)


```

## Ques 7

Below is the R chunk to explain the index working :

Difference : (6) searches the films table by the index of TitleIndex where as
(4) uses primary key (rowid) to search

How to identify : The search from index is specified by index name, here in case
of (6) we have "SEARCH f USING INDEX TitleIndex (title=?)"

Comments : where there is a search by title query having an index in title
column of the table ensures the db engine improve the efficiency of data 
retrieval operations by a sorted copy of the indexed column(s) and 
corresponding data rows.


## QUES 8
We can see that after creating index the operation happened relatively fast,
as it incorporated efficient search with sorted index rather than linear scan
```{r}

time <- system.time({
  film_info <- dbGetQuery(dbcon, "
    SELECT f.title, c.name AS category, f.length
    FROM film f
    JOIN film_category fc ON f.film_id = fc.film_id
    JOIN category c ON fc.category_id = c.category_id
    WHERE f.title = 'ZORRO ARK'
  ")
})



print(time)
  
  
```


## QUES 9
```{r}

# as per readings using lower case because MySQL queries are not case-sensitive  
# by default. If this works different it might be because of the internals of
# engine

query9 <-  "
  SELECT f.title , f.length , l.name
  FROM film as f
  JOIN language as l ON f.language_id = l.language_id
  where f.title LIKE '%gold%'
  "

result9 <- dbGetQuery(dbcon, query9)

print(result9)
  
  
```


```{r}

# as per readings using lower case because MySQL queries are not case-sensitive  
# by default. If this works different it might be because of the internals of
# engine

query10 <-  "
  EXPLAIN QUERY PLAN
  SELECT f.title , f.length , l.name
  FROM film as f
  JOIN language as l ON f.language_id = l.language_id
  where f.title LIKE '%gold%'
  "

result10 <- dbGetQuery(dbcon, query10)

print(result10)

  
  
```
The query doesnt use indexes because :
Indexes are not useful when pattern matching searches are performed with LIKE.
And in the case of searches using LIKE theengines do linear scan anyways.

