---
output:
  pdf_document: default
  html_document: default
---
## Question 4
Load the XML with the DTD into R using validation.
```{r}
library("XML")

xml_data <- xmlParse("CustomersAndOrders.xml",validate = T)

```

## Question 5
Execute an XPath expression that returns the names of all customers that do not live in "MA"
```{r}
cust_name <- xpathApply(xml_data,"/Root/Customers/Customer[FullAddress/Region 
                        != 'MA']/CompanyName",xmlValue)

customer_names <- unlist(cust_name)

print(customer_names)
```

## Question 6
Using the result returned in (5), count how many customers there are.
```{r}
print(length(customer_names))
```

## Question 7
Using a combination of R and XPath, calculate the total amount paid for freight for all orders placed by customer with ID "GREAL".
```{r}
freight_values <- as.numeric(xpathApply(xml_data, "/Root/Orders
                                        /Order[CustomerID = 'GREAL']
                                        /ShipInfo/Freight" , xmlValue))


freight_values <- unlist(freight_values)

print(freight_values)


total_freight_amount <- sum(freight_values)

print(total_freight_amount)
```

## Question 8
Using a combination of R and XPath, calculate the average amount paid for freight for all orders shipped to the USA.
```{r}
a <- as.numeric( xpathApply(xml_data,"/Root/Orders
                            /Order[ShipInfo/ShipCountry='USA']
                            /ShipInfo/Freight",xmlValue) )

a <- unlist(a)

print(mean(a))

```

